---
title: "Converting SPSS Syntax to R Markdown"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Converting SPSS Syntax to R Markdown}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

The `spss2rmarkdown` package converts SPSS `.sps` syntax files into
self-contained R Markdown reports. The generated reports use the **jmv**
package for statistical analyses and preserve original SPSS syntax as
foldable comments for easy comparison.

## The Conversion Pipeline

Conversion follows three steps:

1. **Parse** the `.sps` file into structured command objects.
2. **Convert** each SPSS command to equivalent R code.
3. **Generate** an R Markdown document assembling the converted code.

### Step 1: Parse the Syntax File

`parse_sps()` reads a `.sps` file, strips comments, normalises whitespace,
splits commands on the SPSS period terminator, and returns a list of parsed
command objects.

```{r parse-example, eval = FALSE}
library(spss2rmarkdown)

parsed <- parse_sps("analysis.sps")
length(parsed)
parsed[[1]]$command_type
parsed[[1]]$raw
```

Each element contains fields such as `command_type`, `raw`, `subcommands`,
`variables`, and `options`.

### Step 1b: Parse the Data File

`parse_sav()` reads an SPSS `.sav` file and extracts the data, variable
metadata, and value labels.

```{r parse-sav-example, eval = FALSE}
sav_info <- parse_sav("data.sav")
sav_info$n_obs
sav_info$n_vars
head(sav_info$metadata)
```

You can also generate R code that recreates the data-loading step:

```{r load-code, eval = FALSE}
cat(generate_data_load_code(sav_info, new_path = "data.sav"))
```

### Step 2: Convert

`convert_all_commands()` dispatches each parsed command to the appropriate
converter and returns R code along with metadata.

```{r convert-example, eval = FALSE}
converted <- convert_all_commands(parsed, sav_info)
converted[[1]]$r_code
converted[[1]]$analysis_type
```

For individual commands:

```{r convert-single, eval = FALSE}
result <- convert_spss_to_r(parsed[[1]], sav_info)
cat(result$r_code)
```

### Step 3: Generate R Markdown

`generate_rmd()` assembles a complete `.Rmd` file with YAML header, setup
chunk, data loading, transformations, and analysis sections.

```{r generate-example, eval = FALSE}
rmd <- generate_rmd(
  sav_data       = sav_info,
  sav_path       = "data.sav",
  parsed_syntax  = parsed,
  converted_code = converted,
  output_dir     = tempdir(),
  base_name      = "analysis"
)
rmd$rmd_path
```

## Expression Conversion

The expression converter translates SPSS syntax to R:

```{r expr-examples}
library(spss2rmarkdown)

convert_spss_expression("MEAN(var1, var2, var3)")
convert_spss_expression("MISSING(age)")
convert_spss_expression("x = 1 AND y <> 0")
```

## Full Pipeline Example

```{r full-pipeline, eval = FALSE}
library(spss2rmarkdown)

# Parse
parsed   <- parse_sps("study.sps")
sav_info <- parse_sav("study.sav")

# Expand TO syntax (e.g., item1 TO item10)
parsed <- lapply(parsed, expand_spss_variables, sav_info = sav_info)

# Convert
converted <- convert_all_commands(parsed, sav_info)

# Generate R Markdown
rmd <- generate_rmd(
  sav_data       = sav_info,
  sav_path       = "study.sav",
  parsed_syntax  = parsed,
  converted_code = converted,
  output_dir     = "output"
)

cat("Report written to:", rmd$rmd_path)
```
